Verified Nanopasses for Compiling Conditionals

### 2025-06-18 paper.tex
progress

Olivier Danvy, in his paper \emph{A New One-Pass Transformation into
Monadic Normal Form}~\citep{Danvy:2003fk}, presents a beautiful family
of structurally recursive functions that translate Boolean expressions
and conditionals into a monadic normal form that no longer contains
Boolean expressions or nested conditionals, and where jumps are
expressed using zero-arity function calls. One of the key parts of
that translation is in the compilation of a conditional expression
such as $(\IF{b}{e_1}{e_2})$, where the expression $b$ is compiled
using a special function $\mathcal{B}$ that takes two more arguments,
the generated code for the two branches. Thus, $\mathcal{B}$ can
analyze the expression $b$ to decide how to generate code for the
$\mathsf{if}$, or whether an $\mathsf{if}$ is even necessary.
\[
\CE{\IF{b}{e_1}{e_2}} = \CB{b}{\CE{e_1}}{\CE{e_2}}
\]
For example, in the simple cases where $b$ is $\TRUE$ or $\FALSE$,
no branching is needed.
\begin{align*}
\CB{\TRUE}{c_1}{c_2} &= c_1 \\
\CB{\FALSE}{c_1}{c_2} &= c_2
\end{align*}

My first exposure to these ideas were in Kent Dybvig's compiler
course~\citep{Dybvig:2010aa} where he taught the techniques used in
the Chez Scheme compiler~\citep{Dybvig:2006aa}.  One of those
techniques is called destintation-driven code
generation~\cite{Dybvig:1990aa}, which employs an extra parameter on
the compilation function that indicates how the context will use the
result value of an expression. The context might (1) ignore the result
(and only care about the expression's side effects), (2) branch based
on interpreting its result as a boolean, or (3) actually use the
result value. These options are encoded as a disjoint union.  Relating
this to Olivier's transformation functions, the $\mathcal{E}$ function
is for contexts that need the actual value and the $\mathcal{B}$
function is for contexts that branch on the expression's result. Of
course, a pair of functions is isomorphic to a function on a disjoint
union.
\[
(A \to C) \times (B \to C) \cong (A + B) \to C
\]

The CompCert compiler \citep{Leroy:2006fe}, a compiler for C verified
in Coq, uses the one-function per kind of context approach, in the
RTLgen pass that translates from the CminorSel intermediate language
to its register transfer language.

>>>>

Olivier Danvy, in his paper \emph{A New One-Pass Transformation into
Monadic Normal Form}~\citep{Danvy:2003fk}, presents a beautiful family
of structurally recursive functions that translate Boolean expressions
and conditionals into a monadic normal form that (1) no longer
contains Boolean expressions, (2) the condition expression of a
condition is a value, and (3) jumps are expressed using zero-arity
function calls. One of the key parts of that translation is in the
compilation of a conditional expression such as $(\IF{b}{e_1}{e_2})$,
where the expression $b$ is compiled using a special function
$\mathcal{B}$ that takes two more arguments, the generated code for
the two branches. Thus, $\mathcal{B}$ can analyze the expression $b$
to decide how to generate code for the $\mathsf{if}$, or whether an
$\mathsf{if}$ is even necessary.
\[
\CE{\IF{b}{e_1}{e_2}} = \CB{b}{\CE{e_1}}{\CE{e_2}}
\]
For example, in the simple cases where $b$ is $\TRUE$ or $\FALSE$,
no branching is needed.
\begin{align*}
\CB{\TRUE}{c_1}{c_2} &= c_1 \\
\CB{\FALSE}{c_1}{c_2} &= c_2
\end{align*}

My first exposure to these ideas were in Kent Dybvig's compiler
course~\citep{Dybvig:2010aa} where he taught the techniques used in
the Chez Scheme compiler~\citep{Dybvig:2006aa}.  One of those
techniques is called destintation-driven code
generation~\cite{Dybvig:1990aa}, which adds an extra parameter to the
compilation function to indicate how the context will use the result
value of an expression. The context may (1) ignore the result (and
only care about the expression's side effects), (2) branch based on
interpreting its result as a boolean, (3) assign the result to a
variable, or (4) output the result. These options are encoded as a
disjoint union.  Relating this to Olivier's transformation functions,
the $\mathcal{E}$ function is for contexts that need the actual value
and the $\mathcal{B}$ function is for contexts that branch on the
expression's result. Of course, a pair of functions is isomorphic to a
function on a disjoint union.
\[
(A \to C) \times (B \to C) \cong (A + B) \to C
\]

The CompCert compiler \citep{Leroy:2006fe} (a compiler for C verified
in Coq) uses the one-function per context approach of Danvy in the
RTLgen pass that translates from the CminorSel intermediate language
to CompCert's register transfer language. The verified CakeML
compiler, on the other hand, employs a naive translation of Boolean
expressions and conditionals for the compilation of stackLang the
intermediate language into labLang~\citep{Kumar:2014aa}.

Towards developing a course on verified compilation, I am interested
in adapting verification techniques from CompCert to the compiler
presented in the \emph{Essentials of Compilation}
textbook~\citep{Siek:2023tr,Siek:2023ue}. That book takes an
incremental approach to teaching compilers, that is, it guides the
students through the creation of multiple compilers for a sequence of
languages, each of which adds a language feature to the previous
langauge. For the language with Boolean expressions and conditionals,
the Explicate Control pass of the compiler uses the one-function per
context approach of Danvy to translate into an x86-flavored register
transfer language.  \emph{Essentials of Compilation} also employs the
nanopass approach, breaking up the compiler into many small passes
that are easier for student to digest~\citep{Sarkar:2004fk}.

This paper presents a proof of correctness in Agda for four nanopasses
that translate a source language with let binding, integer arithmetic,
conditional expressions and Booleans into an x86-flavored register
transfer language. These four passes correspond to the
following three passes in \emph{Essentials of Compilation}:
\begin{enumerate}
\item Remove Complex Operands
\item Explicate Control
\item Select Instructions
\end{enumerate}
The process of proving of correctness of the Explicate Control pass
made it clear to me that it has two distinct reponsibilities that
deserve to be separated:
\begin{itemize}
\item Replacing \texttt{let} expressions with assignment statements.
\item Translating Boolean expressions and conditionals into
  \texttt{goto}-based control-flow.
\end{itemize}
(The RTLgen pass of CompCert also handles the above two
responsibilities.)  So in this paper, the Explicate Control pass is
split into two passes, the first named Lift Locals and the second that
keeps the name Explicate Control.
\begin{enumerate}
\item Remove Complex Operands
\item Lift Locals
\item Explicate Control
\item Select Instructions
\end{enumerate}

%% differs in that I also use a big-step semantics for the register
%% transfer language (instead of small step), which streamlines some of
%% the reasoning.
